%!TEX root = /Users/markelikalderon/Documents/Git/formwithoutmatter/aristotle.tex
\chapter*{Acknowledgements} % (fold)
\markboth{\MakeUppercase{Acknowledgements}}{}
\addcontentsline{toc}{chapter}{Acknowledgements}
\label{cha:acknowledgements}

% Having to write a paper, I felt compelled to read the literature. Growing bored, I turned to history. Though it would be churlish of me to name them, I am indebted to these authors for without them the present work would likely not have been written.
% 
% Thanks to the London Underground for providing an environment where much of the reading for the present work was done.

Material that was to become chapters 1 and 3 were presented in talks given at Interuniversity Centre, Dubrovnik, and at UCL in 2011. The material from chapter 6 was presented at the Jowett Society, Oxford University, University of Warwick, and University of Aberdeen in 2011. Thanks to all who participated. Special thanks are due to Ursula Coope whose question on the occasion of presenting at UCL prompted me to write the material of chapter 6.1.4.

I would also like to thank the members of a UCL postgraduate research seminar where a draft of this material was presented. I especially benefitted from the comments of Emily Crampton, Margaret Hampson, and Maarten Steenhagen.

Thanks to Peter Momtchiloff for his help in seeing the present essay into print. Thanks to two anonymous referees who read the manuscript for OUP. The comments provided greatly improved the present work. Craig French generously read the penultimate draft and provided extensive comments for which I am most grateful.

Special thanks are due to Fiona Leigh for the kindly forebearance shown to a senior colleague.

% chapter acknowledgements (end)